---
title: "Simulating the Central Limit Theorem with Exponential Distribution"
author: "Bharat Dogra"
date: "2025-03-24"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Overview
This project investigates the Central Limit Theorem (CLT) using simulations with the exponential distribution. We simulate averages of 40 exponential variables over 1000 trials and compare the resulting sampling distribution with theoretical expectations. We analyze its mean, variance, and shape to confirm convergence to normality. We also conduct basic inferential analysis using the ToothGrowth dataset.
## Simulations
We simulate 1000 averages of 40 exponential random variables with a rate parameter `lambda = 0.2`. The theoretical mean and standard deviation of the exponential distribution are both `1 / lambda = 5`.

```{r simulate-setup}
set.seed(1234)
lambda <- 0.2
n <- 40
num_sim <- 1000

# Simulate sample means
means <- replicate(num_sim, mean(rexp(n, lambda)))
```
## Visualizing the Sampling Distribution
We create a histogram of the sample means and compare the sample mean to the theoretical mean.

```{r sample-mean-check}
# Histogram of sample means
hist(means, main = "Sampling Distribution of the Mean (n = 40)", 
     xlab = "Sample Mean", col = "lightblue", breaks = 40)

# Compare sample mean and theoretical mean
sample_mean <- mean(means)
theoretical_mean <- 1 / lambda
```

## Sample Variance vs Theoretical Variance
We now compare the sample variance of the means to the theoretical variance of the sampling distribution of the mean. The theoretical variance is calculated as `(1 / lambda)^2 / n`.

```{r variance-check-1}
# Calculate sample and theoretical variance
sample_variance <- var(means)
theoretical_variance <- (1 / lambda)^2 / n # Variance of the sample mean

# Output both values
sample_variance
theoretical_variance
```

## Demonstrating Approximate Normality
To show the distribution of sample means is approximately normal, we overlay a normal curve on the histogram and create a Q-Q plot.

```{r normality-check-1, fig.height=4, fig.width=7}
```
# Histogram with normal curve
hist(means, probability = TRUE, 
     main = "Histogram with Normal Curve", 
     col = "lightblue", xlab = "Sample Mean")
curve(dnorm(x, mean = mean(means), sd = sd(means)), 
      col = "red", lwd = 2, add = TRUE)

# Q-Q plot
qqnorm(means)
qqline(means, col = "red", lwd = 2)
```
## ToothGrowth Data Analysis
We now perform basic inferential analysis using the `ToothGrowth` dataset available in R.  
We explore the effect of supplement type and dose on tooth length.

```{r load-toothgrowth-1}
data("ToothGrowth")
ToothGrowth$dose <- as.factor(ToothGrowth$dose)

head(ToothGrowth)
summary(ToothGrowth)
```
# Exploratory Boxplot
library(ggplot2)
ggplot(ToothGrowth, aes(x = dose, y = len, fill = supp)) +
  geom_boxplot() +
  labs(title = "Tooth Length by Supplement Type and Dose",
       x = "Dose (mg/day)", y = "Tooth Length") +
  theme_minimal()
# T-test: Supplement type comparison
t.test(len ~ supp, data = ToothGrowth, var.equal = TRUE)

# T-test: Dose comparisons
# 0.5 vs 1
t.test(len ~ dose, data = subset(ToothGrowth, dose %in% c("0.5", "1")))

# 1 vs 2
t.test(len ~ dose, data = subset(ToothGrowth, dose %in% c("1", "2")))

# 0.5 vs 2
t.test(len ~ dose, data = subset(ToothGrowth, dose %in% c("0.5", "2")))

## Sample Variance vs Theoretical Variance
We now compare the sample variance of the means to the theoretical variance of the sampling distribution of the mean. The theoretical variance is calculated as `(1 / lambda)^2 / n`.

```{r variance-check-2}
# Calculate sample and theoretical variance
sample_variance <- var(means)
theoretical_variance <- (1 / lambda)^2 / n  # Variance of the sample mean

# Output both values
sample_variance
theoretical_variance
```

## Demonstrating Approximate Normality
To show that the distribution of sample means is approximately normal, we overlay a normal curve on the histogram and create a Q-Q plot.

```{r normality-check-2, fig.height=4, fig.width=7}
```
# Histogram with normal curve
hist(means, probability = TRUE,
     main = "Histogram with Normal Curve",
     col = "lightblue", xlab = "Sample Mean")
curve(dnorm(x, mean = mean(means), sd = sd(means)),
      col = "red", lwd = 2, add = TRUE)

# Q-Q plot
qqnorm(means)
qqline(means, col = "red", lwd = 2)
```
## ToothGrowth Data Analysis
We now perform basic inferential analysis using the `ToothGrowth` dataset available in R.  We explore the effect of supplement type and dose on tooth length.

```{r load-toothgrowth-2}
# Load the dataset and prepare it
data("ToothGrowth")
ToothGrowth$dose <- as.factor(ToothGrowth$dose)

# View dataset structure
head(ToothGrowth)
summary(ToothGrowth)
```
# Exploratory Plot
## We create a boxplot to visualize the effect of supplement type and dose on tooth length.


library(ggplot2)
ggplot(ToothGrowth, aes(x = dose, y = len, fill = supp)) +
  geom_boxplot() +
  labs(title = "Tooth Length by Supplement Type and Dose",
       x = "Dose (mg/day)", y = "Tooth Length") +
  theme_minimal()
# Hypothesis Tests
We compare tooth length by supplement type and between dose levels using t-tests.

# Supplement type comparison
t.test(len ~ supp, data = ToothGrowth, var.equal = TRUE)

# Dose level comparisons
t.test(len ~ dose, data = subset(ToothGrowth, dose %in% c("0.5", "1")))
t.test(len ~ dose, data = subset(ToothGrowth, dose %in% c("1", "2")))
t.test(len ~ dose, data = subset(ToothGrowth, dose %in% c("0.5", "2")))
```




